---
title: "DS6372-Applied Statistics-Project2"
author: "Karthik Radhakrishnan"
date: "2024-12-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exploratory Data Analysis

```{r}
# Loading the data

bank <- read.csv("bank.csv",sep=";",header=TRUE,stringsAsFactors = TRUE)
head(bank)
#summary(bank)
str(bank)

# Find if any of the column has missing values
sum(is.na(bank))

```

### Analyzing the distribution of numeric variables
```{r,message=FALSE}
# Numerical variables
numeric_vars <- c("age", "balance", "day","duration", "campaign", "pdays", "previous")
summary(bank[numeric_vars])

# Histograms for numerical variables
#install.packages("ggthemes")
library(ggthemes)
library(ggplot2)
for (var in numeric_vars) {
  print(ggplot(bank, aes_string(var)) + geom_histogram() + ggtitle(paste("Histogram of", var))+theme_economist())
}
```

### Analyzing the distribution of categorical variables

```{r,message=FALSE}
categorical_vars <- c("job", "marital", "education", "default", "housing", "loan", "contact", "month", "poutcome")
for (var in categorical_vars) {
  print(
    ggplot(bank, aes_string(var)) + 
      geom_bar() + 
      ggtitle(paste("Bar plot of", var)) +
      theme_minimal()
  )
}
```

### Creating a pair plot for numeric features to assess the class overlap. 

```{r,message=FALSE}
library(GGally)

numeric_data <- bank[, numeric_vars]

# Create the pair plot
ggpairs(numeric_data, 
        aes(color = bank$y, alpha = 0.7),
        diag = list(continuous = "densityDiag"), # Add density plots on the diagonal
        upper = list(continuous = "cor")) # Add correlation coefficients in the upper panels

```

### Creating loess plots for numeric predictors

```{r}
#converting response variable to numeric and store in a separate variable
bank$y.num <- ifelse(bank$y == "yes",1,0)

#loess plot for all numeric variables

for (var in numeric_vars) {
  print(
    ggplot(bank, aes(x = .data[[var]], y = y.num)) +  # Dynamically map the variable
      geom_point() + 
      geom_smooth(method = "loess", size = 1, span = 1) + 
      ylim(-0.2, 1.2) + 
      ggtitle(paste("Loess plot of", var)) +
      theme_minimal()
  )
}


```



### Creating bar plot for categorical variables

```{r}
library(dplyr)

## Calculating the proportion of term deposit by job
JobSummary <- bank |> 
  group_by(job,y)|> 
  summarise(count=n())|>
  mutate(percent=round(count/sum(count),2))|> 
  arrange((job))

JobSummary

ggplot(JobSummary, aes(x = job, y = percent, fill = y)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of term deposit by job",
    x = "Job",
    y = "Proportion"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1))


## Calculating the proportion of term deposit by marital status
MaritalSummary <- bank |> 
  group_by(marital,y)|> 
  summarise(count=n())|>
  mutate(percent=round(count/sum(count),2))|> 
  arrange((marital))

MaritalSummary

ggplot(MaritalSummary, aes(x = marital, y = percent, fill = y)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of term deposit by marital status",
    x = "Martial Status",
    y = "Proportion"
  ) +
  theme_minimal() 

## Calculating the proportion of term deposit by education
EducationSummary <- bank |> 
  group_by(education,y)|> 
  summarise(count=n())|>
  mutate(percent=round(count/sum(count),2))|> 
  arrange((education))

EducationSummary

ggplot(EducationSummary, aes(x = education, y = percent, fill = y)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of term deposit by education",
    x = "Education",
    y = "Proportion"
  ) +
  theme_minimal()


## Calculating the proportion of term deposit by default
DefaultSummary <- bank |> 
  group_by(default,y)|> 
  summarise(count=n())|>
  mutate(percent=round(count/sum(count),2))|> 
  arrange((default))

DefaultSummary

ggplot(DefaultSummary, aes(x = default, y = percent, fill = y)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of term deposit by default",
    x = "Default",
    y = "Proportion"
  ) +
  theme_minimal() 

## Calculating the proportion of term deposit by housing
HousingSummary <- bank |> 
  group_by(housing,y)|> 
  summarise(count=n())|>
  mutate(percent=round(count/sum(count),2))|> 
  arrange((housing))

HousingSummary

ggplot(HousingSummary, aes(x = housing, y = percent, fill = y)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of term deposit by housing",
    x = "Housing",
    y = "Proportion"
  ) +
  theme_minimal() 


## Calculating the proportion of term deposit by personal loan
LoanSummary <- bank |> 
  group_by(loan,y)|> 
  summarise(count=n())|>
  mutate(percent=round(count/sum(count),2))|> 
  arrange((loan))

LoanSummary

ggplot(LoanSummary, aes(x = loan, y = percent, fill = y)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of term deposit by loan",
    x = "Loan",
    y = "Proportion"
  ) +
  theme_minimal() 

## Calculating the proportion of term deposit by contact type
ContactSummary <- bank |> 
  group_by(contact,y)|> 
  summarise(count=n())|>
  mutate(percent=round(count/sum(count),2))|> 
  arrange((contact))

ContactSummary

ggplot(ContactSummary, aes(x = contact, y = percent, fill = y)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of term deposit by contact",
    x = "Contact",
    y = "Proportion"
  ) +
  theme_minimal() 

## Calculating the proportion of term deposit by last contact month of the year
ContactMonthSummary <- bank |> 
  group_by(month,y)|> 
  summarise(count=n())|>
  mutate(percent=round(count/sum(count),2))|> 
  arrange((month))

ContactMonthSummary

ggplot(ContactMonthSummary, aes(x = month, y = percent, fill = y)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of term deposit by contact month",
    x = "Contact month",
    y = "Proportion"
  ) +
  theme_minimal() 


## Calculating the proportion of term deposit by outcome
OutcomeSummary <- bank |> 
  group_by(poutcome,y)|> 
  summarise(count=n())|>
  mutate(percent=round(count/sum(count),2))|> 
  arrange((poutcome))

OutcomeSummary

ggplot(OutcomeSummary, aes(x = poutcome, y = percent, fill = y)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportion of term deposit by outcome",
    x = "outcome",
    y = "Proportion"
  ) +
  theme_minimal() 

```



### Feature selection for logistic regression model

```{r,message=FALSE}
library(caret)

# train and test validation split
set.seed(1234)

trainIndex<-createDataPartition(bank$y,p=.5,list=FALSE)  #p: proportion of data in train

training<-bank[trainIndex,]
validate<-bank[-trainIndex,]

# fitting a glmnet model with logloss as error metric

fitControl<-trainControl(method="repeatedcv",number=10,repeats=1,classProbs=TRUE, summaryFunction=twoClassSummary)

glmnet.fit <- train(
  y ~ .-y.num,
  data = training,
  method = "glmnet",
  trControl = fitControl,
  tuneGrid = expand.grid(alpha = 1, lambda = seq(0.001, 0.1, length = 10)), # Set alpha = 1
  preProcess = c("center", "scale"),
  metric = "ROC"
)

glmnet.fit

# Extract features selected by Lasso
selected_features <- coef(glmnet.fit$finalModel, glmnet.fit$bestTune$lambda)
print(selected_features)

```

### checking for multicolinearity

```{r}
library(car)

# Fit a logistic regression model
full_model <- glm(y ~ .-y.num, data = training, family = "binomial")

# Calculate VIF for each predictor
vif_values <- vif(full_model)
print(vif_values)

```

### Building a simple Multiple logistic regression model using the values selected by lasso regression
```{r}
logistic.model <- glm(y~job+marital+education+loan+contact+month+duration+previous+poutcome,
                      data=training,
                      family="binomial")

summary(logistic.model)
```

```{r}

# Make predictions on validation set using logistic regression model
pred_prob <- predict(logistic.model, newdata = validate, type = "response")
pred_class <- ifelse(pred_prob > 0.5, "yes", "no")

# Create confusion matrix
library(caret)
conf_matrix <- confusionMatrix(factor(pred_class), factor(validate$y),positive='yes')
print("Confusion Matrix:")
print(conf_matrix)


# Print feature importance
coef_summary <- summary(logistic.model)$coefficients
feature_importance <- data.frame(
  Feature = rownames(coef_summary),
  Coefficient = coef_summary[,1],
  P_value = coef_summary[,4]
)
feature_importance <- feature_importance[order(abs(feature_importance$Coefficient), decreasing = TRUE),]
print("Feature Importance:")
print(feature_importance)
```

### fitting a complex logistic regression model

#### Feature Engineering

```{r}
bank2 <- bank
# Create interactions between significant variables
bank2$duration_success <- bank2$duration * (bank2$poutcome == "success")
bank2$duration_retired <- bank2$duration * (bank2$job == "retired")
bank2$retired_success <- (bank2$job == "retired") * (bank2$poutcome == "success")

# Create binned variables for continuous features
bank2$duration_bin <- cut(bank2$duration, 
                           breaks = quantile(bank2$duration, probs = seq(0, 1, 0.25)),
                           labels = c("short", "medium", "long", "very_long"))

# Create seasonal groupings
bank2$season <- case_when(
  bank2$month %in% c("dec", "jan", "feb") ~ "winter",
  bank2$month %in% c("mar", "apr", "may") ~ "spring",
  bank2$month %in% c("jun", "jul", "aug") ~ "summer",
  TRUE ~ "fall"
)

# Adding Economic Indicators

bank2$age_group <- cut(bank2$age, 
                         breaks = c(0, 30, 45, 60, Inf),
                         labels = c("young", "middle", "senior", "retired"))
bank2$high_balance <- ifelse(bank2$balance > median(bank2$balance), 1, 0)

# Create campaign effectiveness indicator
bank2$campaign_ratio <- bank2$previous / (bank2$campaign + 1)
bank2$recent_contact <- ifelse(bank2$pdays > -1 & bank2$pdays < 30, 1, 0)



```
### fitting Complex logistic regression model. 

```{r}

# train and test validation split
set.seed(1234)

trainIndex2<-createDataPartition(bank2$y,p=.5,list=FALSE)  #p: proportion of data in train

training2<-bank2[trainIndex,]
validate2<-bank2[-trainIndex,]


logistic.complex.model <- glm(
  y ~ job + marital + education + loan + contact + 
      season + duration_bin + poutcome + previous +
      duration_success + duration_retired + retired_success +
      I(duration^2) + job:poutcome + season:contact +
      age_group + high_balance + campaign_ratio +
      age_group:high_balance + 
      poutcome:campaign_ratio,
  data = training2,
  family = "binomial"
)



# Make predictions on validation set using complex logistic regression model
pred_prob <- predict(logistic.complex.model, newdata = validate2, type = "response")
pred_class <- ifelse(pred_prob > 0.5, "yes", "no")

# Create confusion matrix
conf_matrix <- confusionMatrix(factor(pred_class), factor(validate2$y),positive='yes')
print("Confusion Matrix:")
print(conf_matrix)

```
### fitting a LDA Model
```{r}

fitControl<-trainControl(method="repeatedcv",number=10,repeats=1,classProbs=TRUE, summaryFunction=mnLogLoss)

lda.fit<-train(y~age+balance+day+duration+campaign+pdays+previous,
               data=training,
               method="lda",
               trControl=fitControl,
               metric="logLoss")

#Computing predicted probabilities on the validation data using LDA model
lda.predictions <- predict(lda.fit, validate, type = "prob")[,"yes"]

threshold=0.5

lda.class.predictions = as.factor(ifelse(lda.predictions>threshold, "yes", "no"))

confusionMatrix(lda.class.predictions,validate$y,positive="yes")

length(lda.predictions)

# Plot ROC curve
library(pROC)

lda.roc<-roc(response=validate$y,predictor=lda.predictions,levels=c("no","yes"))
plot(lda.roc,print.thres="best")

# Area under the curve for lda model. 

auc(lda.roc)


```

